A comparison of single-cell trajectory inference methods: towards more accurate and robust tools

### 2018-04-05 analysis/paper/paper_latex.Rmd
update paper

Despite their similar overall performance, the topologies predicted by the top methods differed considerably. Trajectories detected by Slingshot and cellTree tended to contain less branches, while those detected by TSCAN and AGA Pseudotime gravitated towards more complex topologies (`r ref("fig", "topology_complexity_comparison")`). AGA Pseudotime for example frequently predicted a tree topology, even when a non-branching topology was present in the data. Indeed, when we assessed how often a method inferred the correct topology, Slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree most frequently predicted the correct tree topology (`r ref("stable", "topology_sensitivity")`). Although the top methods in the ranking were capable at discriminating linear from bifurcating trajectories (with TSCAN predicting the correct topology in respectively `r scales::percent(filter(topology_sensitivity, method_id == 'tscan', trajectory_type == 'directed_linear')$perc_perfect)` and `r scales::percent(filter(topology_sensitivity, method_id == 'tscan', trajectory_type == 'bifurcation')$perc_perfect)` of datasets), the accuracy when more complex tree trajectories were present in the data was still very low, with the best methods, Monocle DDRTree and SCUBA, predicting the correct tree topology in respectively `r scales::percent(filter(topology_sensitivity, method_id == 'mnclddr', trajectory_type == 'rooted_tree')$perc_perfect)` and `r scales::percent(filter(topology_sensitivity, method_id == 'scuba', trajectory_type == 'rooted_tree')$perc_perfect)` of the cases  (`r ref("stable", "topology_sensitivity")`). Inferring the correct topology without parameter tuning is therefore still an open challenge. Currently, inferring a complex trajectory structure still requires a considerable user input, either by setting optimal parameters, or by subjectively choosing a method whose results best fit the user’s expert knowledge.

@@

<!-- A common trajectory analysis framework -->
Critical to the broad application of TI methods is that both the user's input and the method's output is standardised, so that users can easily and quickly try out different methods, compare their output, and run common analysis pipelines such as network inference [@aibar_scenicsinglecellregulatory_2017a] or module detection [@saelens_comprehensiveevaluationmodule_2018]. A common framework to analyse trajectories is especially important when the output of methods contradict, as they often do. We illustrate this in  `r ref("fig", "example_predictions")`. Here, we predicted trajectories on a datasets containing fibroblasts which were reprogrammed towards neurons and myocytes, a bifurcating trajectory [@treutlein_dissectingdirectreprogramming_2016]. The topology of all five methods differs greatly, with Slingshot underestimating the complexity of the topology, while AGA pseudotime and Monocle DDRTree overestimated the complexity, at least compared to what was expected in the data.

```{r echo=FALSE, results="asis"}
add_fig(
  fig_path = figure_file("bifurcating_example.tmp.pdf", experiment_id = "14-example_predictions"),
  ref_id = "example_predictions", 
  caption_main = "An example gold standard and prediction.", 
  caption_text = "As an illustration to our practical guidelines"
)
```

@@

For each dataset we sampled the same number of cells as were present in the reference real dataset, limited to the simulation steps after burn-in. Next, we used the Splatter package [@zappia_splattersimulationsinglecell_2017] to estimate the different characteristics of a real dataset, such as the distributions of average gene expression, library sizes and dropout probabilities. We used Splatter to simulate the expression levels $\lambda_{i,j}$ of housekeeping genes $i$ (to match the number of genes in the reference dataset) in every cell $j$. These were combined with the expression levels of the genes simulated within a trajectory. Next, true counts were simulated using $Y'_{i,j} \sim \text{Poi}(\lambda_{i,j})$. Finally, we simulated dropouts by setting true counts to zero by sampling from a Bernoulli distribution using a dropout probability $\pi^D_{i,j} =\frac{1}{1+e^{-k(\text{ln}(\lambda_{i,j})-x_0)}}$.

@@

```{r}
add_sfig(
  figure_file("edge_flip_comparison.tmp.pdf", experiment_id="3-metric_characterisation/4-edge_flip"), 
  "edge_flip_comparison", 
  paste0("Edge flip score values for common trajectory topologies"), 
  15, 
  5
)
```

>>>>

Despite their similar overall performance, the topologies predicted by the top methods differed considerably. Trajectories detected by Slingshot and cellTree tended to contain less branches, while those detected by TSCAN and AGA Pseudotime gravitated towards more complex topologies (`r ref("fig", "topology_complexity_comparison")`). AGA Pseudotime for example frequently predicted a tree topology, even when a non-branching topology was present in the data. Indeed, when we assessed how often a method inferred the correct topology, Slingshot and TSCAN were good at correctly predicting linear, bifurcating and converging trajectories, while Monocle DDRTree most frequently predicted the correct tree topology (`r ref("stable", "topology_sensitivity")`). Although the top methods in the ranking were capable at discriminating linear from bifurcating trajectories (with TSCAN predicting the correct topology in respectively `r scales::percent(filter(topology_sensitivity, method_id == 'tscan', trajectory_type == 'directed_linear')$perc_perfect)` and `r scales::percent(filter(topology_sensitivity, method_id == 'tscan', trajectory_type == 'bifurcation')$perc_perfect)` of datasets), the accuracy when more complex tree trajectories were present in the data was still very low, with the best methods, Monocle DDRTree and SCUBA, predicting the correct tree topology in respectively `r scales::percent(filter(topology_sensitivity, method_id == 'mnclddr', trajectory_type == 'rooted_tree')$perc_perfect)` and `r scales::percent(filter(topology_sensitivity, method_id == 'scuba', trajectory_type == 'rooted_tree')$perc_perfect)` of the cases  (`r ref("stable", "topology_sensitivity")`). Inferring the correct topology without parameter tuning is therefore still an open challenge.

@@

<!-- A common trajectory analysis framework -->
Critical to the broad applicability of TI methods is standardisation of the input and output interfaces of TI methods, so that users can effortlessly execute TI methods on their dataset of interest, compare different predicted trajectories, and apply downstream analyses such as network inference [@aibar_scenicsinglecellregulatory_2017a] or module detection [@saelens_comprehensiveevaluationmodule_2018]. Our framework is an initial attempt at tackling this problem, and we illustrate its usefulness here by comparing the predicted trajectories of several top-performing methods on a randomly picked dataset (`r ref("fig", "example_predictions")`). In the future, this framework should be extended to allow easier downstream analyses and also to allow easier interpretation by overlaying custom values such as cell type or expression values. In addition, further debate is required in order to arrive at a consensus concerning a common interface for trajectory models.

```{r echo=FALSE, results="asis"}
add_fig(
  fig_path = figure_file("bifurcating_example.tmp.pdf", experiment_id = "14-example_predictions"),
  ref_id = "example_predictions", 
  caption_main = "Demonstration of how a common framework for TI methods facilitates broad applicability using an example dataset.", 
  caption_text = "a) The dataset consists of fibroblasts being reprogrammed towards neurons and myocytes, and thus contains a bifurcating dynamic process  [@treutlein_dissectingdirectreprogramming_2016]. b) A common visualisation of the predicted trajectories allows to easily compare the similarities and differences between predictions. The predictions are in line with earlier results in that Slingshot underestimates the complexity of the topology, AGA Pseudotime and Monocle DDRTree overestimate the complexity, and TSCAN is able to correctly identify the trajectory topology but obtains a suboptimal ordering of the cells. c) More insight can be gained into which cells ended up on which branch by placing the trajectory segments one after the other. d) Method-specific visualisations provide insight into the inner workings of a method's methodology and into how the trajectory was constructed. e) While only TSCAN and cellTree correctly predict the topology, the ordering of Slingshot and AGA Pseudotime on this dataset corresponds better to the expected ordering."
)
```

@@

For each dataset we sampled the same number of cells as were present in the reference real dataset, limited to the simulation steps after burn-in. Next, we used the Splatter package [@zappia_splattersimulationsinglecell_2017] to estimate the different characteristics of a real dataset, such as the distributions of average gene expression, library sizes and dropout probabilities. We used Splatter to simulate the expression levels $\lambda_{i,j}$ of housekeeping genes $i$ (to match the number of genes in the reference dataset) in every cell $j$. These were combined with the expression levels of the genes simulated within a trajectory. Next, true counts were simulated using $Y'_{i,j} \sim \text{Poisson}(\lambda_{i,j})$. Finally, we simulated dropouts by setting true counts to zero by sampling from a Bernoulli distribution using a dropout probability $\pi^D_{i,j} =\frac{1}{1+e^{-k(\text{ln}(\lambda_{i,j})-x_0)}}$.

@@

```{r}
add_sfig(
  figure_file("edge_flip_comparison.tmp.pdf", experiment_id="3-metric_characterisation/4-edge_flip"), 
  "edge_flip_comparison", 
 "Edge flip score values for common trajectory topologies",
  ""
)
```
